Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-03 Manuscript/manuscript.tex
Corrections to sections A-C in Results (i.e. until Affinity and binding headline)

\begin{abstract}
  % insert abstract here
  The binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  consistently determined with different experimental and theoretical methods
  despite their significance and biological relevance in cellular membranes. 
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved with an implicit inclusion of the electronic polarizability
in the polar region of phospholipids,
which we apply according to the 
electronic continuum correction (ECC)~\cite{leontyev11}.
Such an approach has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
We use the the Lipid14 model \cite{dickson14} as a starting point,
because it provided one of the best available descriptions of cation binding \cite{catte16}.
The developed ECC-lipid model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly in the description of the membrane binding affinities to cations and induced
structural effects of sodium and calcium ions.

@@

The same systems were also simulated with the ECC-lipid model for POPC 
and SPC/E model for water using the same setup.
In these simulations ECC was also applied to the cationic surfactant 
in the same way as to POPC;
all charges were scaled with the same factor as for ECC-lipids, i.e. $f_q=0.8$, 
and the same atom types with reduced $\sigma$ parameters from ECC-lipids were used. 
\todo{Provide the parameteres for ECC-surfactant somewere.}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: The chemical structure of POPC and the labeling of the carbon segments.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipid models
    compared with the experimental values from \cite{ferreira13}.
    The size of the points for the head group order parameters correspond 
    the error bars ($\pm 0.02$ for experiments \cite{botan15,ollila16} and ?? for
    simulations {\bf Add the size of the error bars in simulations.}).
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipid models compared with experiments~\cite{Kucerka2011}.
  } 
\end{figure}

\begin{table}
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipid model, and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

The ECC-lipid and Lipid14 models reproduce the experimental x-ray scattering form factors
of POPC bilayer with the similar accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid values from the Lipid14 model is slightly larger than the
experimental value in Table~\ref{tab:apls}, while the value from ECC-lipid model
is slightly smaller. Area per lipid values of the ECC-lipid model show some variation
when simulated with different water models in Table~\ref{tab:apls_si} (Supplementary Information),
however, all the values are close to the experimentally reported values.
In conclusion, the ECC-lipid model reproduces the experimental dimensions of POPC
lipid bilayer with the accuracy comparable to the other state of the art lipid models \cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

The acyl chain order parameters of the Lipid14 model~\cite{dickson14} and the
ECC-lipid model agree with the experimental values with in the error bars
in Fig.~\ref{simVSexpNOions}, althought the ECC-lipid model gives slightly larger
values for {\it sn}-1 chain. Notably, the experimentally measured forking and
small order parameter values of $C_2$ segment in {\it sn}-2 chain are relatively well
reproduced by the both models. This has been suggested to indicate that the carbonyl
of {\it sn}-2 chain is directed towards the water phase, in contrast to the
carbonyl in {\it sn}-1 chain, which would orient more along the bilayer
plane~\cite{seelig75,schindler75,gawrisch92}. While this may be an important
feature for the ion binding details, it is not necessarily reproduced by the
available lipid models~\cite{ollila16}
%This is expected as the ECC model does not modify
%the already highly optimized parameters of this region in Lipid14.

The order parameters of $\alpha$ and $\beta$ carbons in the headgroup are slightly larger
in the ECC-lipid model than in the Lipid14 model, which is apparently related to the
P-N vector orienting 7$^{\circ}$ more toward the water phase in the ECC-lipid model, 
see Fig.~\ref{OrderParameterCHANGESsurf}. With the current data we cannot,
however, conclude which one of the models give the more realistic
headgroup conformations. The ECC-lipid model gives
the $\beta$ carbon order parameter value closer to experiments, while
value for $\alpha$ carbon is better in the Lipid14 model.
Despite of some deviations from the experimental order parameter values
in Fig. \ref{simVSexpNOions},
the accuracy of the both models in the glycerol backbone region
is comparable to the other state
of art lipid models available in literature \cite{botan15}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}

@@

The changes of the headgroup order parameters with an increasing amount of 
the cationic surfactant is compared between experiments~\cite{scherer89} and
simulations in Fig. \ref{OrderParameterCHANGESsurf}.
Approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ},
is observed in simulations and experiments at least for the mole fractions
below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that 
the head group order parameters respond too sensitively to the bound positive charge.
The slope in the ECC-lipid model is in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.

The headgroup P-N vector angle with respect to the membrane normal
is also shown in Fig.~\ref{OrderParameterCHANGESsurf}
as a function of the mole fraction of the cationic surfactant.
As suggested previously~\cite{seelig87}, the headgroup orients
more towards the water phase with the increasing amount of positive
charge in a PC lipid bilayer. The effect is more pronounced in the
Lipid14 model, for which the addition of 50\% mole fraction of the cationic
surfactant leads to the decrease of 20$^{\circ}$ of the P-N vector angle,
while the corresponding change in the ECC-lipid model is 11$^{\circ}$.
The difference is in line with the smaller order parameter 
changes and the reduced charge--dipole interactions in the latter model.
The lesser sensitivity of the P-N vector angle response in the ECC-lipid
model can be considered to be more realistic, because the changes of the headgroup
order parameters as a function of the bound positive charge
are in better agreement with experiments in this model.
The results also imply that the validation and improvements of MD simulation
models are generally required for the reliable simulation studies of the
lipid headgroup responses to ions or other biomolecules, as also concluded
previously \cite{botan15}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    The changes of the head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  \todo{Add temperatures for POPC and DPPC experiments} \\
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of POPC bilayer as a function of NaCl concentrations
    from simulations with different force fields together with experimental data for DPPC~\cite{akutsu81} and POPC~\cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
  \todo{Add temperatures to the experimental data.} \\
\end{figure}

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments by using the electrometer concept, because 
the order parameters decrease proportionally to the amount of bound positive charge~\cite{seelig87,catte16}. 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate the Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduce
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions,
as seen from the data replotted from Ref.~\citenum{catte16} in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}.
The model reproduced the experimentally measured small order parameter changes with NaCl
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ 
concentration were overestimated by the same combination of force field parameters.

Since significant artefacts are reported in simulations with \AA{}qvist ions is water \cite{auffinger07},
we also simulated the Lipid14 model with the ion models by Dang et al.~\cite{smith94,chang1999,dang2006} and
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, Pluharova2014} having
more realistic bulk behaviour. Instead of the improvement in the binding behaviour,
we observed overbinding also of Na$^+$ with these ions models, while the results
with Ca$^{2+}$ were barely affected,
%for the CaCl$_2$ interactions with a POPC bilayer described by the Lipid14 model,
as seen in Figs. \ref{fig:delta_ordPar_CaCl} and S??.%\ref{fig:ordPars_waterModels} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results are in line with the previous work \cite{catte16},
suggesting that the improvements in the lipid parameters are
required to correctly describe the divalent cation binding to PC lipid bilayers.

%To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
%polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
The results from the simulations with the ECC-lipid and the ECC-ion models \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
exhibit an improved behaviour of cation binding to a POPC bilayer in Fig.~\ref{fig:delta_ordPar_CaCl},
showing a good agreement with experiments in the changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations.
Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipid model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.
Also the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was
signifincantly overestimated in the original Lipid14 model, while the ECC-lipid
gives a good agreement with exepriments, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
Also the changes of the P-N vector angle are more prononced
for the Lipid14 model, for which the maximum CaCl$_2$ induced tilting 
toward water phase is approximately 17$^{\circ}$, while the corresponding
value for ECC-lipid model is 6$^{\circ}$.

Furthermore, the overestimated lipid headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models considered in the previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

The ion binding affinity of ECC-lipid model with different water models is compared
in SI. The performance with all water models is better than in in the original Lipid14
parameteres, but the order parameter changes are slightly overestimated
with all four-site water models and with TIP3P.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-lipid simulation.
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}
  SAMULI: I am not sure if this would be needed anymore.}
\end{figure}

>>>>

\begin{abstract}
  % insert abstract here
  The binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  consistently determined with different experimental and theoretical methods
  despite their significance and biological relevance in cellular membranes. 
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipids model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipids model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved with an implicit inclusion of the electronic polarizability
in the polar region of phospholipids,
which we apply according to the 
electronic continuum correction (ECC)~\cite{leontyev11}.
Such an approach has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
We use the the Lipid14 model \cite{dickson14} as a starting point,
because it provided one of the best available descriptions of cation binding \cite{catte16}.
The developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly in the description of the membrane binding affinities to cations and induced
structural effects of sodium and calcium ions.

@@

The same systems were also simulated with the ECC-lipids model for POPC 
and SPC/E model for water using the same setup.
In these simulations ECC was also applied to the cationic surfactant 
in the same way as to POPC;
all charges were scaled with the same factor as for ECC-lipids, i.e. $f_q=0.8$, 
and the same atom types with reduced $\sigma$ parameters from ECC-lipids were used. 
\todo{Provide the parameteres for ECC-surfactant somewere.}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{Kucerka2011}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with the experimental values from \cite{ferreira13}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

\begin{table}
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

The ECC-lipids and Lipid14 models reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a similar accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid values from the Lipid14 model is slightly larger than the
experimental value in Table~\ref{tab:apls}, while the value from ECC-lipids model
is slightly smaller. The values of the area per lipid of the ECC-lipids model show some variation
when simulated with different water models in Table~\ref{tab:apls_si} (Supplementary Information),
however, all the values are close to the experimentally reported values.
In conclusion, the ECC-lipids model reproduces the experimental dimensions of POPC
lipid bilayer with the accuracy comparable to other state of the art lipid models \cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

The acyl chain order parameters of the Lipid14 model~\cite{dickson14} and the
ECC-lipids model agree with the experimental values with in the error bars
in Fig.~\ref{simVSexpNOions}, althought the ECC-lipids model gives slightly larger
values for {\it sn}-1 chain. Notably, the experimentally measured forking and
small order parameter values of $C_2$ segment in {\it sn}-2 chain are relatively well
reproduced by both models. This has been suggested to indicate that the carbonyl
of {\it sn}-2 chain is directed towards the water phase, in contrast to the
carbonyl in {\it sn}-1 chain, which would orient more along the bilayer
plane~\cite{seelig75,schindler75,gawrisch92}. While this may be an important
feature for the ion binding details, it is not necessarily reproduced by the
available lipid models~\cite{ollila16}
%This is expected as the ECC model does not modify
%the already highly optimized parameters of this region in Lipid14.

The order parameters of $\alpha$ and $\beta$ carbons in the headgroup are slightly larger
in the ECC-lipids model than in the Lipid14 model, which is apparently related to the
P-N vector orienting 7$^{\circ}$ more toward the water phase in the ECC-lipids model, 
see Fig.~\ref{OrderParameterCHANGESsurf}. With the current data we cannot,
however, conclude which one of the models give the more realistic
headgroup conformations. The ECC-lipids model gives
the $\beta$ carbon order parameter value closer to experiments, while
value for $\alpha$ carbon is better in the Lipid14 model.
Despite some deviations from the experimental order parameter values
in Fig. \ref{simVSexpNOions},
the accuracy of the both models in the glycerol backbone region
is comparable to the other state
of art lipid models available in literature \cite{botan15}.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}

@@

The changes of the headgroup order parameters with an increasing amount of 
the cationic surfactant is compared between experiments~\cite{scherer89} and
simulations in Fig. \ref{OrderParameterCHANGESsurf}.
Approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ},
is observed in simulations and experiments at least for the mole fractions
below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that 
the head group order parameters respond too sensitively to the bound positive charge.
The slope in the ECC-lipids model is in a very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.

The headgroup P-N vector angle with respect to the membrane normal
is also shown in Fig.~\ref{OrderParameterCHANGESsurf}
as a function of the mole fraction of the cationic surfactant.
As suggested previously~\cite{seelig87}, the headgroup orients
more towards the water phase with the increasing amount of positive
charge in a PC lipid bilayer. The effect is more pronounced in the
Lipid14 model, for which the addition of 50\% mole fraction of the cationic
surfactant leads to the decrease of 20$^{\circ}$ of the P-N vector angle,
while the corresponding change in the ECC-lipids model is 11$^{\circ}$.
The difference is in line with the smaller order parameter 
changes and the reduced charge--dipole interactions in the latter model.
The lower sensitivity of the P-N vector angle response in the ECC-lipid
model can be considered to be more realistic, because the changes of the headgroup
order parameters as a function of the bound positive charge
are in better agreement with experiments in this model.
The results also imply that the validation and improvements of MD simulation
models are generally required for the reliable simulation studies of the
lipid headgroup responses to ions or other biomolecules, as also concluded
previously \cite{botan15}.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    The changes of the head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  \todo{Add temperatures for POPC and DPPC experiments} \\
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of POPC bilayer as a function of NaCl concentrations
    from simulations with different force fields together with experimental data for DPPC~\cite{akutsu81} and POPC~\cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
  \todo{Add temperatures to the experimental data.} \\
\end{figure}

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments using the electrometer concept, because 
the order parameters decrease proportionally 
to the amount of bound positive charge (Fig.~\ref{OrderParameterCHANGESsurf} and Ref.~\citenum{seelig87,catte16}). 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduces
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception in Ref.~\citenum{catte16} was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions.
Such a combination of models reproduced the experimentally measured 
small order parameter changes with NaCl (Fig.~\ref{fig:delta_ordPar_NaCl})
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ concentration 
(Fig.~\ref{fig:delta_ordPar_CaCl}) were overestimated by the same combination of force field parameters.

Since significant artefacts are reported in simulations with \AA{}qvist ions is water \cite{auffinger07},
we also simulated the Lipid14 model with more realistic ion models by Dang et al.~\cite{smith94,chang1999,dang2006}, and
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, Pluharova2014}. 
Instead of an improvement in the binding behaviour,
we observed greatly overestimated binding for \ce{Ca^{2+}} and also for Na$^+$ with such ions models 
(Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}). 
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results are in line with the previous work \cite{catte16},
suggesting that the improvements in the lipid parameters are
required to correctly describe the binding of divalent cations to phospholipid bilayers.

%To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
%polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
The results from the simulations with the ECC-lipids and the ECC-ion models \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
exhibit an improved behaviour of cation binding to a POPC bilayer in Fig.~\ref{fig:delta_ordPar_CaCl},
showing a good agreement with experiments in the changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations.
Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipids model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.

In addition, the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was
signifincantly overestimated in the original Lipid14 model, while the ECC-lipids model
gives a greatly improved agreement with exepriments, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
Also the changes of the P-N vector angle are more pronounced
for the Lipid14 model, for which the largest tilting toward water phase 
induced by a $780\,\mathrm{mM}$ CaCl$_2$ concentration
is approximately 17$^{\circ}$, while the corresponding
value for ECC-lipids model is 6$^{\circ}$ ($820\,\mathrm{mM}$ CaCl$_2$).

Furthermore, the overestimated lipid headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models in a previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

The ion binding affinity of ECC-lipids model with different water models is compared in SI. 
In general, the order parameter changes are slightly overestimated
with the four-site water models, and with TIP3p.
However, the performance of ECC-lipids with any of the water models 
is better than the original Lipid14 model.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-lipids simulation.
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}
  SAMULI: I am not sure if this would be needed anymore.}
\end{figure}
